Dynamics of Laterally Propagating Flames in X-ray Bursts

### 2019-10-13 xrb/paper.tex
do a pass through the intro

X-ray bursts (XRBs) are thermonuclear explosions of an accreted H or
He layer on the surface of a neutron star (see~\citealt{galloway:2017}
for a review).  Observations of bursts can help constrain the
properties of the underlying neutron star, helping to illuminate the
nuclear equation of state~\citep{steiner:2010,ozel:2010}.  Extensive
observations of XRBs have provided evidence that the burning begins in
a localized region and spreads over the surface of the neutron
star~\citep{bhattacharyya:2006,bhattacharyya:2007}.

One dimensional studies of XRBs are very successful in predicting the
lightcurves and recurrance times (see, e.g., \citealt{woosley-xrb}).
These assume spherical symmetry and thus cannot capture the effects of
localized burning spreading across the neutron star.  These studies
have also been used to explore the sensitivity of the burst
observables to accretion and reaction rates~\citep{Jose2010a,Lampe2016}\MarginPar{ref}.

Multidimensional simulations of burning on a neutron star are more
difficult, with both the temporal and spatial scales presenting
challenges~\citep{astronum_2018}.  For the spatial scales, we need to
resolve the reaction zone, $\mathcal{O}(10~\mbox{cm})$ or smaller, the
scale height of the atmosphere, $\mathcal{O}(500~\mbox{cm})$, and the
Rossby scale where the Coriolis force balances the lateral pressure
gradient, $\mathcal{O}(10^5~\mbox{cm})$~\citep{spitkovsky2002}.  In
the temporal scales, capturing the rise, $\mathcal{O}(1~\mbox{s})$,
and the decay of the burst, $\mathcal{O}(10~\mbox{s})$, and the
accretion period between bursts, $\mathcal{O}(10^4~\mbox{s})$ is
beyond the ability of multidimensional hydro codes currently.
Nevertheless, significant progress has been made in understanding the
multidimensional nature of XRBs, through various approximations.x

@@

Small-domain studies of convective burning in XRBs preceding flame
development have been done in two-dimensions \citep{lin:2006,xrb,xrb2}
and three-dimensions \citep{xrb3d}.  These calculations used low Mach
number methods that approximate the hydrodynamics equations to filter
soundwaves, enabling large timesteps and efficient modeling of
subsonic convection.  While these calculations could not support
the lateral differences needed for flame spreading, they can help
understand the role that convection plays in distributing the initial
burning products vertically throughout the neutron star atmosphere.

@@

Detonations are in some sense easier to model, since their structure
follows from the jump conditions which are solved with shock-capturing
schemes.  For deflagrations, we either need to resolve the structure
or use a flame model.  Flame models usually assume that the flame
structure is thin compared to the size of the system (see, e.g., \citet{Ropke2007}\MarginPar{refs} for applications to Type Ia supernovae).  For XRBs however, the flame thickness is comparable to the scale
height of the atmosphere, so we cannot use these approximations.

>>>>

X-ray bursts (XRBs) are thermonuclear explosions of an accreted H or
He layer on the surface of a neutron star (see~\citealt{galloway:2017}
for a review).  Observations of bursts can help constrain the
properties of the underlying neutron star, helping to illuminate the
nuclear equation of state~\citep{steiner:2010,ozel:2010}.  Extensive
observations of brightness oscillations during the rise have provided
evidence that the burning begins in a localized region and spreads
over the surface of the neutron
star~\citep{bhattacharyya:2006,bhattacharyya:2007,chakraborty:2014}.

One dimensional studies of XRBs are very successful in predicting the
lightcurves and recurrance times (see, e.g., \citealt{woosley-xrb}).
These assume spherical symmetry and thus cannot capture the effects of
localized burning spreading across the neutron star.  These studies
have also been used to explore the sensitivity of the burst
observables to accretion and reaction
rates~\citep{cyburt:2010,Jose2010a,Lampe2016}, and model individual
bursts~\citep{johnston:2019}.

Multidimensional simulations of burning on a neutron star are more
difficult, with both the temporal and spatial scales presenting
challenges (see \citealt{astronum_2018} for an overview).  For the spatial scales, we need to
resolve the reaction zone, $\mathcal{O}(10~\mbox{cm})$ or smaller, the
scale height of the atmosphere, $\mathcal{O}(500~\mbox{cm})$, and the
Rossby scale where the Coriolis force balances the lateral pressure
gradient, $\mathcal{O}(10^5~\mbox{cm})$~\citep{spitkovsky2002}.  For
the temporal scales, capturing the rise, $\mathcal{O}(1~\mbox{s})$,
and the decay of the burst, $\mathcal{O}(10~\mbox{s})$, and the
accretion period between bursts, $\mathcal{O}(10^4~\mbox{s})$ is
beyond the ability of multidimensional hydro codes currently.
Nevertheless, significant progress has been made in understanding the
multidimensional nature of XRBs, through various approximations.x

@@

Small-domain studies of convective burning in XRBs preceding flame
development have been done in two-dimensions \citep{lin:2006,xrb,xrb2}
and three-dimensions \citep{xrb3d}.  These calculations used low Mach
number methods that approximate the hydrodynamics equations to filter
soundwaves, enabling large timesteps and efficient modeling of
subsonic convection.  While these calculations could not support
the lateral differences needed for flame spreading, they can help
understand the role that convection plays in distributing the initial
burning products vertically throughout the neutron star atmosphere as well
as the nature of any turbulence the burning front might encounter.

@@

For deflagrations, we either need to resolve the structure of the
reaction zone or use a flame model.  Flame models usually assume that
the flame structure is thin compared to the size of the system (see,
e.g., \citet{Ropke2007} for applications to Type Ia supernovae).  For
XRBs however, the flame thickness is comparable to the scale height of
the atmosphere, so we cannot use these approximations.  Accurate
models of flames therefore require that we resolve the thermal width,
which is $\mathcal{O}(10~cm)$ for helium flames~\citep{Timmes00}.
